# Cognition {.unnumbered}


```{r}
#| warning: false
#| message: false
#| code-fold: true

library(tidyverse)
library(gt)

df = read_csv("https://raw.githubusercontent.com/ghidinelli/fred-jehle-spanish-verbs/master/jehle_verb_database.csv")

my_gt_theme <- function(gt_object, ...) {
  
  stopifnot("'gt_object' must be a 'gt_tbl', have you accidentally passed raw data?" = "gt_tbl" %in% class(gt_object))
  
  tab_out <- gt_object %>%
    # Fonts
    opt_table_font(
      font = list(
        google_font("Noto Sans"), 
        google_font("Source Sans Pro"),
        default_fonts()
      )
    ) %>%
    
    # Cell styles from both themes
    tab_style(
      style = cell_borders(
        sides = "top", color = "white", weight = px(0)),
      locations = cells_body(rows = ifelse(nrow(gt_object[["_data"]]) > 0, 1, NA))) %>%
    tab_style(
      style = cell_text(color = "#005689", size = px(22), weight = 700),
      locations = list(cells_title(groups = "title"))) %>%
    tab_style(
      style = cell_text(color = "#545454", font = google_font("Source Sans Pro"), transform = "uppercase"),
      locations = cells_column_labels(everything())) %>%
    tab_style(
      style = cell_text(font = google_font("Libre Franklin"), weight = 800),
      locations = cells_title(groups = "title")) %>%
    tab_style(
      style = cell_text(font = google_font("Source Sans Pro"), weight = 400),
      locations = cells_body()) %>%
    
    # Table options combining elements of both themes
    tab_options(
      # Table-wide settings
      table.background.color = "white",
      table.font.size = 16,
      
      # Table border settings
      table.border.top.style = "none",
      table.border.top.width = px(1),
      table.border.top.color = "#40c5ff",
      table.border.bottom.width = px(3),
      table.border.bottom.color = "white",
      
      # Column label settings
      column_labels.background.color = "#f6f6f6",
      column_labels.font.weight = "bold",
      column_labels.border.top.style = "none",
      column_labels.border.bottom.style = "none",
      column_labels.border.bottom.width = 1,
      column_labels.border.bottom.color = "#334422",
      
      # Data row settings
      row.striping.include_table_body = TRUE,
      row.striping.background_color = "#ececec",
      data_row.padding = px(6), # Average of both themes
      
      # Row group settings
      row_group.font.weight = "800",
      
      # Heading settings
      heading.align = "left",
      heading.border.bottom.style = "none",
      
      # Table body border settings
      table_body.border.top.style = "none",
      
      # Source notes settings
      source_notes.font.size = 12,
      
    )
  
  return(tab_out)
}

create_conjugation_table <- function(verb_infinitive) {
  
  verb_order = c("Presente", 
                 "Imperfecto", 
                 "Pretérito", 
                 "Futuro", 
                 "Condicional", 
                 "Presente perfecto",
                 "Pluscuamperfecto",
                 "Pretérito anterior",
                 "Futuro perfecto",
                 "Condicional perfecto")
  
  verb_df <- df %>%
    filter(infinitive == verb_infinitive)
  
  infinitive = verb_df %>% pull(infinitive) %>% .[1] 
  english_infinitive = verb_df %>% pull(infinitive_english) %>% .[1] %>% str_to_title
  gerund = verb_df %>% pull(gerund) %>% .[1] 
  past_participle = verb_df %>% pull(pastparticiple) %>% .[1] 
  
  verb_table <- verb_df %>%
    select(-c(tense_english, 
              infinitive, 
              infinitive_english, 
              verb_english,
              # mood, 
              mood_english,
              gerund, 
              gerund_english, 
              pastparticiple, 
              pastparticiple_english
              )) %>%
    mutate(
      temp_form_2p = form_2p,
      form_2p = ifelse(mood %in% c("Imperativo Afirmativo","Imperativo Negativo"), form_3s, form_2p),
      form_3s = ifelse(mood %in% c("Imperativo Afirmativo","Imperativo Negativo"), temp_form_2p, form_3s),
      form_1p = ifelse(is.na(form_1p), paste0(form_3s, "mos"), form_1p),
      form_1s = ifelse(is.na(form_1s), " - ", form_1s)
    ) %>%
    select(-temp_form_2p) %>% 
    mutate(tense = factor(tense, levels = verb_order)) %>% 
    arrange(tense) %>% 
    group_by(mood) %>% 
    gt() %>%
    tab_header(
      title = paste("Conjugation of",infinitive),
      subtitle = md(glue::glue(sprintf("English: *{english_infinitive}*\n
                                     Gerundio: *{gerund}*\n
                                     Pasado Participio: *{past_participle}*")))) %>%
    cols_label(
      tense = "Tense",
      # verb_english = "English",
      form_1s = "Yo",
      form_2s = "Tú",
      form_3s = "Él/Ella/Usted",
      form_1p = "Nosotros",
      form_2p = "Vosotros",
      form_3p = "Ellos/Ellas/Ustedes"
    ) %>% 
    cols_align(
      align = c("right"),
      columns = everything()
    ) %>% 
    cols_align(
      align = c("left"),
      columns = tense
    ) %>% 
    my_gt_theme()

  return(verb_table)
}

```


::: {.panel-tabset}


# pensar

## to think

We think that pensar is a very useful verb for you to learn. It is used in a great number of idiomatic expressions and everyday situations. Pay attention to the stem change in Tenses 1 and 6 and in the imperative mood.

Pensar is used with the preposition de when you ask someone what he/she thinks of someone or something, i.e., when you ask for someone’s opinion: 

 - ¿Qué piensa Ud. de este libro?
   - What do you think of this book? 
 - Pienso que es bueno
   - I think that it is good. 
   
Pensar is used with the preposition en when you ask someone what or whom he/she is thinking about:

  - Miguel, no hablas mucho; ¿en qué piensas?
    - Michael, you are not talking much; what are you thinking of? (of what are you thinking?); 
  - Pienso en las vacaciones de verano
    - I’m thinking of summer vacation.


## Conjugation Table

```{r}
create_conjugation_table("pensar")
```


## Example Phrases

- **¿Qué piensa Ud. de eso?**
  - What do you think of that?
- **¿En qué piensa Ud.?**
  - What are you thinking of?
- **Pensamos comprar una casa.**
  - We intend to buy a house.

## Proverbios

- **Antes de hablar es bueno pensar.**
  - (It’s good to) think before you speak.

## Dichos

- **pensativo, pensativa**
  - thoughtful, pensive
- **un pensador, una pensadora**
  - a thinker
- **pensar + inf.**
  - to intend + inf.
- **pensar en**
  - to think of, about
- **sin pensar**
  - thoughtlessly
- **repensar**
  - to think over (again)
- **pensante**
  - thinking
- **pensativamente**
  - thoughtfully
- **el pensamiento**
  - thought
- **estar absorto en sus pensamientos**
  - to be deep in thought


# esperar

## to expect, to hope, to wait (for)

## Conjugation Table

```{r}
create_conjugation_table("esperar")
```


## Example Phrases

- **Mientras hay vida hay esperanza.**
  - When there is life, there is hope.
- **la esperanza**
  - hope
- **No hay esperanza.**
  - There is no hope.
- **Espere, por favor.**
  - Please wait.
- **dar esperanzas**
  - to give encouragement
- **desesperar**
  - to despair
- **la espera**
  - wait, waiting
- **la sala de espera**
  - waiting room


# creer


## to believe 


We believe that creer is a useful verb for you to learn because it has an important stem change in Tenses 3 and 7. This verb is helpful in a great number of everyday expressions and situations.

## Conjugation Table

```{r}
create_conjugation_table("creer")
```


## Example Phrases

- **Hasta que no lo veas, no lo creas.**
  - Don’t believe it until you see it.
- **Ver es creer.**
  - Seeing is believing.
- **¡Ya lo creo!**
  - Of course!
- **Creo que sí.**
  - I think so.
- **Creo que no.**
  - I don’t think so.
- **No me lo creo.**
  - I can’t believe it.
- **Se cree buen cantante, pero yo creo que él sería mejor como mimo.**
  - He thinks he’s a good singer, but I believe that he would be better as a mime.

## Dichos

- **crédulo, crédula**
  - credulous, gullible
- **la credulidad**
  - credulity
- **tener buenas creederas**
  - to be credulous, very gullible
- **crédulamente**
  - credulously
- **el credo**
  - creed
- **dar crédito**
  - to believe
- **descreer**
  - to disbelieve
- **incrédulo, incrédula**
  - incredulous
- **la incredulidad**
  - incredulity, unbelief, disbelief
- **increíble**
  - incredible
- **creíble**
  - believable
- **creerse**
  - to consider, to think, to believe (oneself)


# saber

## to know, to know how 

Saber is a very useful irregular verb for you to know. It is used in a great number of idiomatic expressions and everyday situations. Don’t confuse saber with conocer

**Generally speaking,** *saber* means to know a fact, to know something thoroughly:

- **¿Sabe Ud. qué hora es?**/Do you know what time it is?
- **¿Sabe Ud. la lección?**/Do you know the lesson?

When you use *saber + inf.*, it means to know how:

- **¿Sabe Ud. nadar?**/Do you know how to swim?
- **Sí, (yo) sé nadar**/Yes, I know how to swim.

In the preterit tense, *saber* means to find out:

- **¿Lo sabe Ud.?**/Do you know it?
- **Sí, lo supe ayer**/Yes, I found it out yesterday.



## Conjugation Table

```{r}
create_conjugation_table("saber")
```


## Example Phrases

- **Este niño no sabe contar.**
  - This child can’t (does not know how to) count.
- **No sé nada de este asunto.**
  - I don’t know anything about this matter.
- **¿Sabe usted si hay una farmacia por aquí?**
  - Do you know if there is a pharmacy around here?

## Proverbios

- **Saber es poder.**
  - Knowledge is power.
- **Más vale saber que haber.**
  - Knowing is better than having. (Knowledge is worth more than possessions/things.)

## Dichos

- **sabio, sabia**
  - wise, learned
- **un sabidillo, una sabidilla**
  - a know-it-all
- **un sabihondo, una sabihonda**
  - a know-it-all
- **un sabelotodo**
  - a know-it-all
- **la sabiduría**
  - knowledge, learning, wisdom
- **Que yo sepa...**
  - As far as I know...
- **¡Quién sabe!**
  - Who knows! Perhaps! Maybe!
- **la señorita Sabelotodo**
  - Miss Know-It-All
- **el señor Sabelotodo**
  - Mr. Know-It-All
- **sabido, sabida**
  - known
- **el saber**
  - knowledge
- **saber por experiencia**
  - to learn from experience
- **saber algo como el avemaría**
  - to know something like the back of one’s hand
- **no saber ni jota**
  - to have no idea, to not know at all
- **sabiamente**
  - learnedly, wisely




# conocer

Conocer is a verb that you should know! Pay special attention to the spelling change from c to zc before a or o. Conocer is useful in numerous everyday expressions and situations 

Be careful when choosing between **conocer** and **saber**, which also means to know.

- **(a)** Generally speaking, *conocer* means to know in the sense of being acquainted with a person, a place, or a thing:
  - **¿Conoce Ud. a María?**  
    Do you know Mary?
  - **¿Conoce Ud. bien los Estados Unidos?**  
    Do you know the United States well?
  - **¿Conoce Ud. este libro?**  
    Do you know this book? (Are you acquainted with this book?)

In the preterit tense, *conocer* means met in the sense of first met, first became acquainted with someone:
  - **¿Conoce Ud. a Elena?**  
    Do you know Helen?
  - **Sí, (yo) la conocí anoche en casa de un amigo mío.**  
    Yes, I met her (for the first time) last night at the home of one of my friends.

Generally speaking, *saber* means to know a fact, to know something thoroughly:
  - **¿Sabe Ud. qué hora es?**  
    Do you know what time it is?  
    (See *saber* for further examples.)


## Conjugation Table

```{r}
create_conjugation_table("conocer")
```


## Example Phrases

- **El árbol se conoce por su fruta.**
  - The tree is known by its fruit. (You will be judged by your actions; the apple doesn’t fall far from the tree.)
- **–¿Conoce Ud. a esa mujer? –Sí, la conozco.**
  - –Do you know that woman? –Yes, I know her.

## Dichos

- **un conocido, una conocida**
  - an acquaintance
- **conocido, conocida**
  - known
- **conocidamente**
  - obviously
- **desconocido, desconocida**
  - unknown
- **un desconocido, una desconocida**
  - stranger, someone you don’t know
- **un conocedor, una conocedora**
  - an expert
- **el conocimiento**
  - knowledge
- **el desconocimiento**
  - ignorance
- **poner en conocimiento de**
  - to inform (about)
- **muy conocido**
  - very well-known
- **reconocible**
  - recognizable
- **reconocido, reconocida**
  - grateful



::: {.callout-note}

If you can conjugate *conocer*, you can also conjugate the following verbs:

- **reconocer**
  - to recognize, to admit
- **desconocer**
  - to be ignorant of

:::

# parecer

## to seem, to appear

## Conjugation Table

```{r}
create_conjugation_table("parecer")
```

## Example Phrases

- **a lo que parece**
  - according to what it seems
- **al parecer**
  - seemingly, apparently
- **pareciente**
  - similar
- **parecidamente**
  - similarly
- **María parece contenta.**
  - Maria seems happy.
- **Me parece . . .**
  - It seems to me . . .
- **por el bien parecer**
  - for the sake of appearances
- **parecer cansado (cansada)**
  - to look (seem) tired
- **Se parecen como dos gotas de agua.**
  - They resemble each other like two drops of water (like two peas in a pod).
- **parecido, parecida**
  - similar
- **el parecido**
  - resemblance

# suponer

## to suppose, to assume

## Conjugation Table

```{r}
create_conjugation_table("suponer")
```

- **un suponer, una suposición**
  - supposition
- **poner**
  - to put
- **proponer**
  - to propose
- **por supuesto**
  - of course

# sentir/sentirse

## to feel sorry, to regret, to feel / to feel (well, ill), to suffer 

Sentir and sentirse are an essential pair of -ir verbs for you to learn. They are used in a great number of idiomatic expressions and everyday situations. Pay attention to the stem change in certain tenses, as well as in the imperative and present participle (Gerundio).

## Conjugation Table

```{r}
create_conjugation_table("sentir")
create_conjugation_table("sentirse")
```


## Example Phrases

- **La semana pasada sentimos un fuerte terremoto.**
  - Last week we felt a powerful earthquake.
- **El chico sintió la muerte de su perro.**
  - The boy felt very sorry about the death of his dog.
- **Lo siento mucho.**
  - I regret it very much. I’m very sorry.
- **¿Cómo se siente Ud.?**
  - How do you feel?
- **Me siento mal.**
  - I feel sick.

## Dichos

- **el sentimiento**
  - feeling, sentiment
- **sentimentalmente**
  - sentimentally
- **el sentir**
  - feeling, judgment
- **el sentido**
  - sense, meaning, feeling
- **los sentidos**
  - the senses
- **una persona sentimental**
  - sentimentalist
- **el sentimentalismo**
  - sentimentalism
- **sentir admiración por alguien**
  - to feel admiration for someone
- **sentir en el alma**
  - to feel/regret deeply
- **sentido, sentida**
  - sincere, sensitive
- **resentirse**
  - to feel the effects
- **resentirse de algo**
  - to resent something
- **sentirse mal**
  - to feel bad


:::