# Learning {.unnumbered}


```{r}
#| warning: false
#| message: false
#| code-fold: true

library(tidyverse)
library(gt)

df = read_csv("https://raw.githubusercontent.com/ghidinelli/fred-jehle-spanish-verbs/master/jehle_verb_database.csv")

my_gt_theme <- function(gt_object, ...) {
  
  stopifnot("'gt_object' must be a 'gt_tbl', have you accidentally passed raw data?" = "gt_tbl" %in% class(gt_object))
  
  tab_out <- gt_object %>%
    # Fonts
    opt_table_font(
      font = list(
        google_font("Noto Sans"), 
        google_font("Source Sans Pro"),
        default_fonts()
      )
    ) %>%
    
    # Cell styles from both themes
    tab_style(
      style = cell_borders(
        sides = "top", color = "white", weight = px(0)),
      locations = cells_body(rows = ifelse(nrow(gt_object[["_data"]]) > 0, 1, NA))) %>%
    tab_style(
      style = cell_text(color = "#005689", size = px(22), weight = 700),
      locations = list(cells_title(groups = "title"))) %>%
    tab_style(
      style = cell_text(color = "#545454", font = google_font("Source Sans Pro"), transform = "uppercase"),
      locations = cells_column_labels(everything())) %>%
    tab_style(
      style = cell_text(font = google_font("Libre Franklin"), weight = 800),
      locations = cells_title(groups = "title")) %>%
    tab_style(
      style = cell_text(font = google_font("Source Sans Pro"), weight = 400),
      locations = cells_body()) %>%
    
    # Table options combining elements of both themes
    tab_options(
      # Table-wide settings
      table.background.color = "white",
      table.font.size = 16,
      
      # Table border settings
      table.border.top.style = "none",
      table.border.top.width = px(1),
      table.border.top.color = "#40c5ff",
      table.border.bottom.width = px(3),
      table.border.bottom.color = "white",
      
      # Column label settings
      column_labels.background.color = "#f6f6f6",
      column_labels.font.weight = "bold",
      column_labels.border.top.style = "none",
      column_labels.border.bottom.style = "none",
      column_labels.border.bottom.width = 1,
      column_labels.border.bottom.color = "#334422",
      
      # Data row settings
      row.striping.include_table_body = TRUE,
      row.striping.background_color = "#ececec",
      data_row.padding = px(6), # Average of both themes
      
      # Row group settings
      row_group.font.weight = "800",
      
      # Heading settings
      heading.align = "left",
      heading.border.bottom.style = "none",
      
      # Table body border settings
      table_body.border.top.style = "none",
      
      # Source notes settings
      source_notes.font.size = 12,
      
    )
  
  return(tab_out)
}

create_conjugation_table <- function(verb_infinitive) {
  
  verb_order = c("Presente", 
                 "Imperfecto", 
                 "Pretérito", 
                 "Futuro", 
                 "Condicional", 
                 "Presente perfecto",
                 "Pluscuamperfecto",
                 "Pretérito anterior",
                 "Futuro perfecto",
                 "Condicional perfecto")
  
  verb_df <- df %>%
    filter(infinitive == verb_infinitive)
  
  infinitive = verb_df %>% pull(infinitive) %>% .[1] 
  english_infinitive = verb_df %>% pull(infinitive_english) %>% .[1] %>% str_to_title
  gerund = verb_df %>% pull(gerund) %>% .[1] 
  past_participle = verb_df %>% pull(pastparticiple) %>% .[1] 
  
  verb_table <- verb_df %>%
    select(-c(tense_english, 
              infinitive, 
              infinitive_english, 
              verb_english,
              # mood, 
              mood_english,
              gerund, 
              gerund_english, 
              pastparticiple, 
              pastparticiple_english
              )) %>%
    mutate(
      temp_form_2p = form_2p,
      form_2p = ifelse(mood %in% c("Imperativo Afirmativo","Imperativo Negativo"), form_3s, form_2p),
      form_3s = ifelse(mood %in% c("Imperativo Afirmativo","Imperativo Negativo"), temp_form_2p, form_3s),
      form_1p = ifelse(is.na(form_1p), paste0(form_3s, "mos"), form_1p),
      form_1s = ifelse(is.na(form_1s), " - ", form_1s)
    ) %>%
    select(-temp_form_2p) %>% 
    mutate(tense = factor(tense, levels = verb_order)) %>% 
    arrange(tense) %>% 
    group_by(mood) %>% 
    gt() %>%
    tab_header(
      title = paste("Conjugation of",infinitive),
      subtitle = md(glue::glue(sprintf("English: *{english_infinitive}*\n
                                     Gerundio: *{gerund}*\n
                                     Pasado Participio: *{past_participle}*")))) %>%
    cols_label(
      tense = "Tense",
      # verb_english = "English",
      form_1s = "Yo",
      form_2s = "Tú",
      form_3s = "Él/Ella/Usted",
      form_1p = "Nosotros",
      form_2p = "Vosotros",
      form_3p = "Ellos/Ellas/Ustedes"
    ) %>% 
    cols_align(
      align = c("right"),
      columns = everything()
    ) %>% 
    cols_align(
      align = c("left"),
      columns = tense
    ) %>% 
    my_gt_theme()

  return(verb_table)
}

```


::: {.panel-tabset}

# escribir

# estudiar

# leer

# entender

# aprender

# decir (to say, to tell)

Decir is an extremely useful irregular verb. You can use decir in many everyday situations and idiomatic expressions.

## Conjugation Table

```{r}
create_conjugation_table("decir")
```

## Example Phrases

- **Dígame, por favor, dónde está la sala de espera.**
  - Tell me, please, where the waiting room is located.
- **¿Qué quiere decir esta palabra?**
  - What does this word mean?
  - **¡Te he dicho mil veces de ponerte el cinturón de seguridad!**
  - I’ve told you a thousand times to put on your seat belt!
  
  ## Proverbios
  
  - **Dime con quién andas y te diré quién eres.**
  - Tell me who your friends are and I will tell you who you are.
- **Al decir las verdades se pierdan las amistades.**
  - Friendships are lost when the truth is told.
- **De decir a hacer hay mucho que ver.**
  - There is a great difference between saying and doing.

## Dichos

- **Dicho y hecho.**
  - No sooner said than done.
- **querer decir**
  - to mean
- **un decir**
  - a familiar saying
- **Diga or Dígame**
  - Hello (on the telephone)
- **decirle al oído**
  - to whisper in one’s ear
- **no decir palabra**
  - not to say a word
- **Es decir...**
  - That is to say...
- **un dicho**
  - a saying, expression
- **dictar**
  - to dictate
- **el dictado**
  - dictation
- **el dictador**
  - dictator
- **decir mentiras**
  - to tell lies
- **decir adiós**
  - to say good-bye
- **bendecir**
  - to bless
- **maldecir**
  - to curse, to swear




# hablar

# llamar/llamarse

:::