# Movement A {.unnumbered}


```{r}
#| warning: false
#| message: false
#| code-fold: true

library(tidyverse)
library(gt)

df = read_csv("https://raw.githubusercontent.com/ghidinelli/fred-jehle-spanish-verbs/master/jehle_verb_database.csv")

my_gt_theme <- function(gt_object, ...) {
  
  stopifnot("'gt_object' must be a 'gt_tbl', have you accidentally passed raw data?" = "gt_tbl" %in% class(gt_object))
  
  tab_out <- gt_object %>%
    # Fonts
    opt_table_font(
      font = list(
        google_font("Noto Sans"), 
        google_font("Source Sans Pro"),
        default_fonts()
      )
    ) %>%
    
    # Cell styles from both themes
    tab_style(
      style = cell_borders(
        sides = "top", color = "white", weight = px(0)),
      locations = cells_body(rows = ifelse(nrow(gt_object[["_data"]]) > 0, 1, NA))) %>%
    tab_style(
      style = cell_text(color = "#005689", size = px(22), weight = 700),
      locations = list(cells_title(groups = "title"))) %>%
    tab_style(
      style = cell_text(color = "#545454", font = google_font("Source Sans Pro"), transform = "uppercase"),
      locations = cells_column_labels(everything())) %>%
    tab_style(
      style = cell_text(font = google_font("Libre Franklin"), weight = 800),
      locations = cells_title(groups = "title")) %>%
    tab_style(
      style = cell_text(font = google_font("Source Sans Pro"), weight = 400),
      locations = cells_body()) %>%
    
    # Table options combining elements of both themes
    tab_options(
      # Table-wide settings
      table.background.color = "white",
      table.font.size = 16,
      
      # Table border settings
      table.border.top.style = "none",
      table.border.top.width = px(1),
      table.border.top.color = "#40c5ff",
      table.border.bottom.width = px(3),
      table.border.bottom.color = "white",
      
      # Column label settings
      column_labels.background.color = "#f6f6f6",
      column_labels.font.weight = "bold",
      column_labels.border.top.style = "none",
      column_labels.border.bottom.style = "none",
      column_labels.border.bottom.width = 1,
      column_labels.border.bottom.color = "#334422",
      
      # Data row settings
      row.striping.include_table_body = TRUE,
      row.striping.background_color = "#ececec",
      data_row.padding = px(6), # Average of both themes
      
      # Row group settings
      row_group.font.weight = "800",
      
      # Heading settings
      heading.align = "left",
      heading.border.bottom.style = "none",
      
      # Table body border settings
      table_body.border.top.style = "none",
      
      # Source notes settings
      source_notes.font.size = 12,
      
    )
  
  return(tab_out)
}

create_conjugation_table <- function(verb_infinitive) {
  
  verb_order = c("Presente", 
                 "Imperfecto", 
                 "Pretérito", 
                 "Futuro", 
                 "Condicional", 
                 "Presente perfecto",
                 "Pluscuamperfecto",
                 "Pretérito anterior",
                 "Futuro perfecto",
                 "Condicional perfecto")
  
  verb_df <- df %>%
    filter(infinitive == verb_infinitive)
  
  infinitive = verb_df %>% pull(infinitive) %>% .[1] 
  english_infinitive = verb_df %>% pull(infinitive_english) %>% .[1] %>% str_to_title
  gerund = verb_df %>% pull(gerund) %>% .[1] 
  past_participle = verb_df %>% pull(pastparticiple) %>% .[1] 
  
  verb_table <- verb_df %>%
    select(-c(tense_english, 
              infinitive, 
              infinitive_english, 
              verb_english,
              # mood, 
              mood_english,
              gerund, 
              gerund_english, 
              pastparticiple, 
              pastparticiple_english
              )) %>%
    mutate(
      temp_form_2p = form_2p,
      form_2p = ifelse(mood %in% c("Imperativo Afirmativo","Imperativo Negativo"), form_3s, form_2p),
      form_3s = ifelse(mood %in% c("Imperativo Afirmativo","Imperativo Negativo"), temp_form_2p, form_3s),
      form_1p = ifelse(is.na(form_1p), paste0(form_3s, "mos"), form_1p),
      form_1s = ifelse(is.na(form_1s), " - ", form_1s)
    ) %>%
    select(-temp_form_2p) %>% 
    mutate(tense = factor(tense, levels = verb_order)) %>% 
    arrange(tense) %>% 
    group_by(mood) %>% 
    gt() %>%
    tab_header(
      title = paste("Conjugation of",infinitive),
      subtitle = md(glue::glue(sprintf("English: *{english_infinitive}*\n
                                     Gerundio: *{gerund}*\n
                                     Pasado Participio: *{past_participle}*")))) %>%
    cols_label(
      tense = "Tense",
      # verb_english = "English",
      form_1s = "Yo",
      form_2s = "Tú",
      form_3s = "Él/Ella/Usted",
      form_1p = "Nosotros",
      form_2p = "Vosotros",
      form_3p = "Ellos/Ellas/Ustedes"
    ) %>% 
    cols_align(
      align = c("right"),
      columns = everything()
    ) %>% 
    cols_align(
      align = c("left"),
      columns = tense
    ) %>% 
    my_gt_theme()

  return(verb_table)
}

```


::: {.panel-tabset}

# entrar

## to enter, to go (in), to come (in) 

## Conjugation Table
  
```{r}
create_conjugation_table("entrar")
```

# llegar

## to arrive, to reach 

## Conjugation Table
  
```{r}
create_conjugation_table("llegar")
```

## Example Sentences

- **Luis y Luisa quieren llegar a ser médicos.**
  - Louis and Louise want to become doctors.

- **Perdimos el avión por haber llegado demasiado tarde al aeropuerto.**
  - We missed the plane because we had arrived too late at the airport.

## Example Phrases

- **llegar a saber**
  - to find out

- **la llegada**
  - arrival

- **llegar tarde**
  - to arrive late

- **llegar a ser**
  - to become

- **llegar a**
  - to reach

- **al llegar**
  - on arrival, upon arriving


# venir 

## to come

Venir is a very important irregular verb for beginning students. It is used in a great number of idiomatic expressions and everyday situations. Pay special attention to the stem changes!
  
## Conjugation Table
  
```{r}
create_conjugation_table("venir")
```

## Example Phrases

- **La semana que viene voy a hacer un viaje.**
  - Next week, I’m going to take a trip.
- **La señora González y su marido vienen de Venezuela.**
  - Mrs. González and her husband come from (are from) Venezuela.
- **Prefiero quedarme en casa porque viene una tormenta.**
  - I’d rather stay home because a storm is coming.

## Dichos

- **la venida**
  - arrival, coming
- **la semana que viene**
  - next week
- **el mes que viene**
  - next month
- **el porvenir**
  - the future
- **en lo porvenir**
  - in the future
- **venidero, venidera**
  - future
- **Venga lo que venga.**
  - Come what may.
- **Viene a ser lo mismo.**
  - It amounts to the same thing.
- **venir a las manos**
  - to come to blows
- **venir a buscar**
  - to come for, to get
- **venir a la mente**
  - to come to mind
- **venir a la cabeza**
  - to come to mind
- **venir a ser lo mismo**
  - to amount to the same thing
- **venir al mundo**
  - to be born, to come into the world
- **venir como anillo al dedo**
  - to be just right, to be fitting (el anillo = ring, el dedo = finger), to fit like a glove
- **bienvenido, bienvenida**
  - welcome
- **dar la bienvenida a**
  - to welcome
- **¡Bienvenido! ¡Bienvenida!**
  - Welcome!
  - **venirse**
  - to come, to come back

# seguir

## to follow, to pursue, to continue

## Conjugation Table
  
```{r}
create_conjugation_table("seguir")
```

## Example Phrases

- **según**
  - according to

- **al día siguiente**
  - on the following day

- **las frases siguientes**
  - the following sentences

- **seguir + pres. part.**
  - to keep on + pres. part.
    - e.g., **Siga leyendo.** - Keep on reading.

- **seguido, seguida**
  - continuous

- **conseguir**
  - to attain, to get, to obtain

- **proseguir**
  - to continue, proceed

- **perseguir**
  - to pursue

- **seguirle los pasos a uno**
  - to keep one’s eye on someone

- **un seguidor, una seguidora**
  - follower


# caminar

## to walk, to move along

## Conjugation Table
  
```{r}
create_conjugation_table("caminar")
```

## Example Phrases

- **el camino**
  - road, highway

- **el camino de hierro**
  - railroad

- **en camino de**
  - on the way to

- **una caminata**
  - a long walk

- **hacer de un camino dos mandados**
  - to kill two birds with one stone

- **el camino real**
  - highway, high road

- **estar en camino**
  - to be on one’s way

- **quedarse a medio camino**
  - to stop halfway

- **por buen camino**
  - on the right road

- **al buen camino**
  - on the right track

- **un camión**
  - truck


# andar

## to walk

Andar is a very useful verb for a beginning Spanish student. Pay special attention to the spelling changes

## Example Phrases

- **¿Cómo andan los negocios?**
  - How’s business?
  
- **Anda despacio que tengo prisa.**
  - Make haste slowly.
  
- **Amadís de Gaula fue un caballero andante de la Edad Media.**
  - Amadis of Gaul was a knight-errant of the Middle Ages.
  
- **¡Anda a pasear!**
  - Take a walk! (Take a hike!)
  
- **¡A Magdalena le gusta andar a caballo! Anda a caballo tres veces por semana.**
  - Madeleine loves horseback riding! She rides a horse three times a week.

## Proverbios

- **Dime con quién andas y te diré quién eres.**
  - Tell me who your friends are and I will tell you who you are.
  
- **Poco a poco se anda lejos.**
  - One step at a time. (Little by little, one goes far away.)

## Dichos

- **andarse**
  - to go away
  
- **las andanzas**
  - events
  
- **buena andanza**
  - good fortune
  
- **mala andanza**
  - bad fortune
  
- **a todo andar**
  - at full speed
  
- **desandar**
  - to retrace one’s steps
  
- **andante**
  - errant
  
- **un caballero andante**
  - knight-errant
  
- **Anda con Dios.**
  - Go with God.
  
- **andar con cien ojos**
  - to be cautious (to have eyes on the back of one’s head)
  
- **el andar**
  - gait (way of walking)
  
- **andar a gatas**
  - to crawl, to walk/on all fours
  
- **andar a caballo**
  - to ride a horse


## Conjugation Table
  
```{r}
create_conjugation_table("andar")
```

# quedar/quedarse

## to be left / to remain, to stay 

Quedarse is a very important reflexive verb for a beginning student. It is used in a great number of idiomatic expressions and everyday situations.

## Conjugation Table
  
```{r}
create_conjugation_table("quedar")
create_conjugation_table("quedarse")
```

## Example Phrases

- **¿Cómo andan los negocios?**
  - How’s business?
  
- **Anda despacio que tengo prisa.**
  - Make haste slowly.
  
- **Amadís de Gaula fue un caballero andante de la Edad Media.**
  - Amadis of Gaul was a knight-errant of the Middle Ages.
  
- **¡Anda a pasear!**
  - Take a walk! (Take a hike!)
  
- **¡A Magdalena le gusta andar a caballo! Anda a caballo tres veces por semana.**
  - Madeleine loves horseback riding! She rides a horse three times a week.

## Proverbios

- **Dime con quién andas y te diré quién eres.**
  - Tell me who your friends are and I will tell you who you are.
  
- **Poco a poco se anda lejos.**
  - One step at a time. (Little by little, one goes far away.)

## Dichos

- **andarse**
  - to go away
  
- **las andanzas**
  - events
  
- **buena andanza**
  - good fortune
  
- **mala andanza**
  - bad fortune
  
- **a todo andar**
  - at full speed
  
- **desandar**
  - to retrace one’s steps
  
- **andante**
  - errant
  
- **un caballero andante**
  - knight-errant
  
- **Anda con Dios.**
  - Go with God.
  
- **andar con cien ojos**
  - to be cautious (to have eyes on the back of one’s head)
  
- **el andar**
  - gait (way of walking)
  
- **andar a gatas**
  - to crawl, to walk/on all fours
  
- **andar a caballo**
  - to ride a horse


# salir

## to go out, to leave

Salir is a very useful irregular verb for you to learn. It is used in a great number of idiomatic expressions and everyday situations.

**Be careful not to confuse _salir_ and _dejar_.** Use _salir de_ when you mean to leave in the sense of to go out of (a place):
> El alumno salió de la sala de clase.  
> The pupil left the classroom.


## Conjugation Table
  
```{r}
create_conjugation_table("salir")
```



## Example Phrases

- **¿A qué hora sale el tren para San José?**
  - At what time does the train leave for San José?

- **—¿Dónde está su madre?**
  - —Where is your mother?
- **—Mi madre salió.**
  - —My mother went out.

## Dichos

- **la salida**
  - exit, departure
  
- **la salida de emergencia**
  - emergency exit
  
- **la salida de incendios**
  - fire exit
  
- **sin salida**
  - no exit, dead-end street
  
- **salir de compras**
  - to go out shopping
  
- **salir mal**
  - to go wrong, to do badly
  
- **saliente**
  - salient, prominent
  
- **salir al encuentro de**
  - to go to meet
  
- **salir de**
  - to leave from, to get out of
  
- **salga lo que salga**
  - come what may
  
- **salir disparado**
  - to take off like a shot, like an arrow
  
- **salir rana**
  - to be a disappointment (una rana: a frog)


# dejar

## to let, to permit, to allow, to leave

## Conjugation Table
  
```{r}
create_conjugation_table("dejar")
```

## Example Phrases

- **dejar caer**
  - to drop (to let fall)

- **el dejo**
  - abandonment

- **dejado, dejada**
  - dejected

- **El alumno dejó sus libros en la sala de clase.**
  - The pupil left his books in the classroom.

- **dejar de + inf.**
  - to stop, to fail to

- **Los alumnos dejaron de hablar cuando la profesora entró en la sala de clase.**
  - The students stopped talking when the teacher came into the classroom.

- **¡No deje Ud. de llamarme!**
  - Don’t fail to call me!

- **dejarse**
  - to abandon (neglect) oneself

- **dejar atrás**
  - to leave behind

- **¡Déjelo!**
  - Let it! (Leave it!)


:::