# Movement B {.unnumbered}


```{r}
#| warning: false
#| message: false
#| code-fold: true

library(tidyverse)
library(gt)

df = read_csv("https://raw.githubusercontent.com/ghidinelli/fred-jehle-spanish-verbs/master/jehle_verb_database.csv")

my_gt_theme <- function(gt_object, ...) {
  
  stopifnot("'gt_object' must be a 'gt_tbl', have you accidentally passed raw data?" = "gt_tbl" %in% class(gt_object))
  
  tab_out <- gt_object %>%
    # Fonts
    opt_table_font(
      font = list(
        google_font("Noto Sans"), 
        google_font("Source Sans Pro"),
        default_fonts()
      )
    ) %>%
    
    # Cell styles from both themes
    tab_style(
      style = cell_borders(
        sides = "top", color = "white", weight = px(0)),
      locations = cells_body(rows = ifelse(nrow(gt_object[["_data"]]) > 0, 1, NA))) %>%
    tab_style(
      style = cell_text(color = "#005689", size = px(22), weight = 700),
      locations = list(cells_title(groups = "title"))) %>%
    tab_style(
      style = cell_text(color = "#545454", font = google_font("Source Sans Pro"), transform = "uppercase"),
      locations = cells_column_labels(everything())) %>%
    tab_style(
      style = cell_text(font = google_font("Libre Franklin"), weight = 800),
      locations = cells_title(groups = "title")) %>%
    tab_style(
      style = cell_text(font = google_font("Source Sans Pro"), weight = 400),
      locations = cells_body()) %>%
    
    # Table options combining elements of both themes
    tab_options(
      # Table-wide settings
      table.background.color = "white",
      table.font.size = 16,
      
      # Table border settings
      table.border.top.style = "none",
      table.border.top.width = px(1),
      table.border.top.color = "#40c5ff",
      table.border.bottom.width = px(3),
      table.border.bottom.color = "white",
      
      # Column label settings
      column_labels.background.color = "#f6f6f6",
      column_labels.font.weight = "bold",
      column_labels.border.top.style = "none",
      column_labels.border.bottom.style = "none",
      column_labels.border.bottom.width = 1,
      column_labels.border.bottom.color = "#334422",
      
      # Data row settings
      row.striping.include_table_body = TRUE,
      row.striping.background_color = "#ececec",
      data_row.padding = px(6), # Average of both themes
      
      # Row group settings
      row_group.font.weight = "800",
      
      # Heading settings
      heading.align = "left",
      heading.border.bottom.style = "none",
      
      # Table body border settings
      table_body.border.top.style = "none",
      
      # Source notes settings
      source_notes.font.size = 12,
      
    )
  
  return(tab_out)
}

create_conjugation_table <- function(verb_infinitive) {
  
  verb_order = c("Presente", 
                 "Imperfecto", 
                 "Pretérito", 
                 "Futuro", 
                 "Condicional", 
                 "Presente perfecto",
                 "Pluscuamperfecto",
                 "Pretérito anterior",
                 "Futuro perfecto",
                 "Condicional perfecto")
  
  verb_df <- df %>%
    filter(infinitive == verb_infinitive)
  
  infinitive = verb_df %>% pull(infinitive) %>% .[1] 
  english_infinitive = verb_df %>% pull(infinitive_english) %>% .[1] %>% str_to_title
  gerund = verb_df %>% pull(gerund) %>% .[1] 
  past_participle = verb_df %>% pull(pastparticiple) %>% .[1] 
  
  verb_table <- verb_df %>%
    select(-c(tense_english, 
              infinitive, 
              infinitive_english, 
              verb_english,
              # mood, 
              mood_english,
              gerund, 
              gerund_english, 
              pastparticiple, 
              pastparticiple_english
              )) %>%
    mutate(
      temp_form_2p = form_2p,
      form_2p = ifelse(mood %in% c("Imperativo Afirmativo","Imperativo Negativo"), form_3s, form_2p),
      form_3s = ifelse(mood %in% c("Imperativo Afirmativo","Imperativo Negativo"), temp_form_2p, form_3s),
      form_1p = ifelse(is.na(form_1p), paste0(form_3s, "mos"), form_1p),
      form_1s = ifelse(is.na(form_1s), " - ", form_1s)
    ) %>%
    select(-temp_form_2p) %>% 
    mutate(tense = factor(tense, levels = verb_order)) %>% 
    arrange(tense) %>% 
    group_by(mood) %>% 
    gt() %>%
    tab_header(
      title = paste("Conjugation of",infinitive),
      subtitle = md(glue::glue(sprintf("English: *{english_infinitive}*\n
                                     Gerundio: *{gerund}*\n
                                     Pasado Participio: *{past_participle}*")))) %>%
    cols_label(
      tense = "Tense",
      # verb_english = "English",
      form_1s = "Yo",
      form_2s = "Tú",
      form_3s = "Él/Ella/Usted",
      form_1p = "Nosotros",
      form_2p = "Vosotros",
      form_3p = "Ellos/Ellas/Ustedes"
    ) %>% 
    cols_align(
      align = c("right"),
      columns = everything()
    ) %>% 
    cols_align(
      align = c("left"),
      columns = tense
    ) %>% 
    my_gt_theme()

  return(verb_table)
}

```


::: {.panel-tabset}

# acabar

## to finish, to end, to complete

Acabar is an important verb to learn because it is a regular -ar verb and because there are many everyday expressions related to it.


## Conjugation Table

```{r}
create_conjugation_table("acabar")
```

### The Spanish Idiomatic Expression

**acabar de + inf.** in Spanish is translated to English as **to have just + past participle**.

#### Present Indicative:

When you use _acabar_ in the present tense, it indicates that the action of the main verb (+ inf.) has just occurred now in the present. In English, we express this using _have just_ + the past participle of the main verb.

**Examples:**

- **María acaba de llegar.**  
  - _Mary has just arrived._
- **Acabo de comer.**  
  - _I have just eaten._
- **Acabamos de terminar la lección.**  
  - _We have just finished the lesson._

#### Imperfect Indicative:

When using _acabar_ in the imperfect indicative, it denotes that the action of the main verb (+ inf.) had occurred at some past time when another action transpired. In English, this is articulated using _had just_ + the past participle of the main verb.

**Examples:**

 - **María acababa de llegar.**  
   - _Mary had just arrived._
 - **Yo acababa de comer.**
   - _I had just eaten._
 - **Acabábamos de terminar la lección.**  
   - We had just finished the lesson.

## Proverbio

 - Bien está lo que bien acaba.
   - All’s well that ends well.


## Dichos

 - **el acabamiento**
   - completion
 - **acabar con**
   - to put an end to
 - **acabar en**
   - to end in
 - **acabar por**
   - to end by, to ... finally
 - **acabado, acabada**
   - finished



::: {.callout-note}

> **Note**: When _acabar_ is used in the imperfect indicative + inf. (of the main verb being expressed), the verb in the other clause is typically in the preterit.  
**Example:**  
Acabábamos de entrar cuando el teléfono sonó.  
_We had just entered the house when the telephone rang._

:::


# comenzar

## to begin, to start, to commence

Comenzar is a useful verb for beginning Spanish students because it has an important stem change and a spelling change from z to c before the letter e. It is also used in numerous everyday expressions and situations.

## Conjugation Table

```{r}
create_conjugation_table("comenzar")
```

## Example Phrases

  - **–¿Qué tiempo hace?**
    - —What is the weather like?
  - **–Comienza a llover.**
    - —It’s starting to rain.
  - **–Quiero comenzar al comienzo.**
    - —I’d like to begin at the beginning.
  - **–¡Comienza!**
    - —Begin!

## Proverb

  - **Lo que mal comienza, mal acaba.**
    - What starts badly ends badly.

## Dichos

  - **el comienzo**
    - the beginning
  - **Note:** la ceremonia de graduación (school)
    - commencement
  - **al comienzo**
    - at the beginning, at first
  - **comenzar a + inf.**
    - to begin + inf.
  - **comenzar por + inf.**
    - to begin by + pres. part.
  - **comenzar por el principio**
    - to begin at the beginning
  - **recomenzar**
    - to begin again


# llevar

## to carry (away), to take (away), to wear

Llevar is an essential regular -ar verb for beginning students of Spanish. It is used in many idiomatic expressions and everyday situations.

**Llevar** generally translates to "to take" in English, especially in the context of carrying or transporting items from one place to another.

- **Example**:  
  José llevó la silla de la cocina al comedor.  
  _Joseph took the chair from the kitchen to the dining room._

Moreover, **llevar** is used when indicating taking someone to a place.

- **Example**:  
  Pedro llevó a María al baile anoche.  
  _Peter took Mary to the dance last night._

Additionally, **llevar** can mean "to wear".

- **Example**:  
  María, ¿por qué llevas la falda nueva?  
  _Mary, why are you wearing your new skirt?_

## Conjugation Table

```{r}
create_conjugation_table("llevar")
```

## Example Phrases

  - **Lo llevo conmigo.**
    - I’ll take it with me.
  - **Aquel hombre lleva una vida de perros.**
    - That man leads a dog’s life.

## Dichos

  - **llevar a cabo**
    - to carry through, to accomplish
  - **llevar una caída**
    - to have a fall
  - **llevar una vida de perros**
    - to lead a dog’s life
  - **llevar conmigo**
    - to take with me
  - **un llevador, una llevadora**
    - carrier
  - **llevarse**
    - to take away, carry off
  - **llevarse algo de alguien**
    - to take something from someone
  - **llevarse bien con alguien**
    - to get along well with someone

# traer

## to bring, to carry, to wear

Traer is a very useful irregular verb for beginning students. It is used in a great number of idiomatic expressions and everyday situations. Pay special attention to the stem changes!

## Conjugation Table

```{r}
create_conjugation_table("traer")
```

## Example Phrases

  - **Tráigame una silla, por favor.**
    - Bring me a chair, please.
  - **¿Qué te trae por aquí?**
    - What brings you here?
  - **¿Quién trajo el pastel?**
    - Who brought the cake?
  - **Yo lo traje.**
    - I brought it.

## Dichos

  - **el traje**
    - costume, dress, suit
  - **el traje de baño**
    - bathing suit
  - **el traje de novia**
    - wedding dress
  - **el traje hecho**
    - ready-made suit
  - **el traje a la medida**
    - tailor-made suit
  - **¡trae! ¡traiga!**
    - Give it here! Give it to me!
  - **traer y llevar**
    - to spread rumors
  - **contraer**
    - to contract
  - **traer entre manos**
    - to have in mind
  - **traer a la mente**
    - to bring to mind
  - **trajear**
    - to clothe
  - **traer buena suerte**
    - to bring good luck
  - **traer mala suerte**
    - to bring bad luck
  - **traer cola**
    - to bring trouble
  - **traerse bien/mal**
    - to dress nicely/poorly


# pasar

## to pass (by), to happen, to spend (time), to go through

## Conjugation Table

```{r}
create_conjugation_table("pasar")
```

## Example Phrases

  - **pasajero, pasajera**
    - passenger, traveler
  - **¡Que lo pase Ud. bien!**
    - Good luck, good-bye!
  - **¿Qué pasa?**
    - What’s happening? What’s going on?
  - **el pasatiempo**
    - amusement, pastime
  - **¿Qué te pasa?**
    - What’s the matter with you?
  - **¡Pase un buen día!**
    - Have a nice day!
  - **pasar un examen**
    - to take an exam

## Note

Use **pasar** when you spend time: *Me gustaría pasar un año en Costa Rica* - I would like to spend a year in Costa Rica. (See **gastar**.)


# acercar/acercarse

## to bring near, to place near / to approach, to draw near

## Conjugation Table

```{r}
create_conjugation_table("acercar")
create_conjugation_table("acercarse")
```

## Example Phrases

  - **acerca de**
    - about, regarding, with regard to
  - **el acercamiento**
    - approaching, approximation
  - **cerca de**
    - near
  - **de cerca**
    - close at hand, closely
  - **acerca de esto**
    - hereof
  - **la cerca**
    - fence, hedge
  - **el cercado**
    - fenced-in area
  - **mis parientes cercanos**
    - my close relatives
  - **cercanamente**
    - soon, shortly
  - **cercano, cercana**
    - near, close
  - **cercar**
    - to enclose, fence in
  - **las cercanías**
    - neighborhood, suburbs


# caer/caerse

## to fall / to fall, to fall down, to tumble

Caer and caerse are an important pair of verbs for a beginning student. Both verbs are irregular and they are very useful in many everyday situations and expressions. Pay special attention when you use caer reflexively (caerse).

## Conjugation Table

```{r}
create_conjugation_table("caer")
```

## Example Phrases

  - **Mi madre cayó enferma en octubre.**
    - My mother fell ill in October.
  - **Yo me caí por la ventana. Por suerte, estuve en la planta baja.**
    - I fell out of the window. Luckily, I was on the first floor.
  - **El hombre que se levanta aún es más grande que el que no ha caído. (Concepción Arenal)**
    - The man who gets up is even greater than the one who hasn’t fallen.

## Proverb

  - **Caen las hojas y caerán los árboles.**
    - All in good time. (Lit.: Leaves fall and trees will fall.)

## Dichos

  - **la caída**
    - the fall
  - **a la caída del sol**
    - at sunset
  - **a la caída de la tarde**
    - at the end of the afternoon
  - **caer enfermo (enferma)**
    - to fall sick
  - **dejar caer**
    - to drop
  - **dejar caer la voz**
    - to drop one’s voice
  - **caer de espaldas**
    - to fall backwards
  - **caer con**
    - to come down with
  - **caer de lo alto**
    - to fall from above
  - **caer de plano**
    - to fall flat
  - **caer en la cuenta**
    - to catch on, to realize, to get the point
  - **caerse a pedazos**
    - to fall to pieces
  - **caerse de risa**
    - to roll (on the floor) with laughter
  - **hacer caer**
    - to knock over

**Similar Conjugations**

  - **decaer**
    - to decay, decline
  - **recaer**
    - to relapse, fall back

# volver

## to return, to go back

Volver is a very useful -er verb for beginning students. It is used in a vast number of idiomatic expressions and everyday situations. Be careful to make the stem change in tenses, as well as the imperative and past participle.

## Conjugation Table

```{r}
create_conjugation_table("volver")
```

## Example Phrases

  - **¿A qué hora vuelve Ud. a casa?**
    - At what time are you going back home?
  - **¿Cuándo podré volver?**
    - When may I return?
  - **Vuelve la página, por favor.**
    - Turn the page, please.

## Dichos

  - **volver en sí**
    - to regain consciousness, to come to
  - **volver sobre sus pasos**
    - to retrace one’s steps
  - **volver atrás**
    - to turn back
  - **una vuelta**
    - a turn, revolution, turning
  - **dar una vuelta**
    - to take a stroll
  - **volverse triste**
    - to become sad
  - **un revólver**
    - revolver, pistol
  - **revolver**
    - to revolve, shake (up), to turn around
  - **volver con las manos vacías**
    - to return empty-handed
  - **volverse**
    - to turn (oneself) around
  - **volverse loco**
    - to go mad
  - **devolver un artículo**
    - to return an article
  - **devolver**
    - to return, go back
  - **envolver**
    - to wrap, to involve, to surround
  - **envolverse**
    - to become involved


# terminar

## to end, to terminate, to finish

## Conjugation Table

```{r}
create_conjugation_table("terminar")
```

## Example Phrases

  - **la terminación**
    - termination, ending, completion
  - **el término**
    - end, ending; term
  - **en otros términos**
    - in other terms, in other words
  - **determinar**
    - to determine
  - **llevar a término**
    - to complete
  - **estar en buenos términos con**
    - to be on good terms with
  - **la terminal aérea**
    - air terminal
  - **el terminal**
    - terminal (electrical or computer)



:::